Using Fundamental Measure Theory to treat the Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2013-08-16 papers/pair-correlation/paper.tex
move references to end of paper in pair paper Tested everything on bingley.
remove unwanted paragraphs around equations Tested everything on darcy.
add sketchy derivation of g_sigma' to pair paper Tested everything on darcy.

\documentclass[letterpaper,twocolumn,amsmath,amssymb,pre]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

@@

The function has the form:

\begin{align}
  g(|\rr|) &= 1 + g_{\sigma}e^{-x_0(r-\sigma)} + x_1g_{\sigma}\sin(x_2(r-\sigma))e^{-x_3(r-\sigma)}\\
  &~~-x_4g_{\sigma}^2\sin(x_5(r-\sigma))e^(-x_6(r-\sigma))
\end{align}

The finalalized form of our pair correlation function is:

@@

%% The homogeneoues limit $g_{sigm}$ can be calculated using the well
%% known and very successful Carnahan Starling equation while the
%% homogeneous limit radial distribution function can be calculated using
%% Monte-Carlo methods.  We used these calculations in a least squares
%% fit along the data points at increasing radial distances in order to
%% fit the perameters of our more general pair correlation function.  Our
%% function will therefore approach the most accurate solution possible
%% in the homogeneous limit.  Created a function that has the form of an
%% expansion in powers of $g_{\sigma}$.  $g_{sigm}$ in the homogeneous
%% limit can be calculated using the well known and very successful
%% Carnahan Starling equation.  has a well known and Perameters in the
%% expansion where fit
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g}\\
  \includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}
\section{What does it look like (graphs that show the function itself)?}
\subsection{Separable fit to the radial distribution function}
Figure~\ref{fig:radial-distribution} shows...

@@

\section{How well does it work?}
Figure~\ref{fig:dadz} shows...
\begin{figure}
  \includegraphics[width=3cm]{figs/dadz-3-2.pdf}
  \includegraphics[width=3cm]{figs/dadz-3-3.pdf}
  \includegraphics[width=3cm]{figs/dadz-2-2.pdf}
  \includegraphics[width=3cm]{figs/dadz-2-3.pdf}
  %%\includegraphics[width=3cm]{figs/dadz-4-3}
  \caption{Plot of $\frac{da_1}{dz}$ near a hard
    wall. For what $\Phi$?}\label{fig:dadz}
\end{figure}

>>>>

\documentclass[letterpaper,twocolumn,amsmath,amssymb,pre,aps,10pt]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

@@

The function has the form:
\begin{align}
  g(|\rr|) &= 1 + g_{\sigma}e^{-x_0(r-\sigma)} + x_1g_{\sigma}\sin(x_2(r-\sigma))e^{-x_3(r-\sigma)}\\
  &~~-x_4g_{\sigma}^2\sin(x_5(r-\sigma))e^(-x_6(r-\sigma))
\end{align}
The finalalized form of our pair correlation function is:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{align}

@@

%% The homogeneoues limit $g_{sigm}$ can be calculated using the well
%% known and very successful Carnahan Starling equation while the
%% homogeneous limit radial distribution function can be calculated using
%% Monte-Carlo methods.  We used these calculations in a least squares
%% fit along the data points at increasing radial distances in order to
%% fit the perameters of our more general pair correlation function.  Our
%% function will therefore approach the most accurate solution possible
%% in the homogeneous limit.  Created a function that has the form of an
%% expansion in powers of $g_{\sigma}$.  $g_{sigm}$ in the homogeneous
%% limit can be calculated using the well known and very successful
%% Carnahan Starling equation.  has a well known and Perameters in the
%% expansion where fit
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}
\section{What does it look like (graphs that show the function itself)?}
\subsection{Separable fit to the radial distribution function}
Figure~\ref{fig:radial-distribution} shows...

@@

\section{How well does it work?}
Figure~\ref{fig:dadz} shows...

\section{Properties of $g_\sigma$}
We begin with some definitions of basic quantities, looking at the
homogeneous hard-sphere fluid.  I am ignoring here the distinction
between $N$ and $N-1$, which is unimportant in the thermodynamic limit
$N\gg 1$.
\begin{align}
  Z(\sigma) &= \iiiint d\rr_1\cdots d\rr_N \prod_{i<j} \Theta(r_{ij}-\sigma)
  \\
  n^{(2)}(\rr_1,\rr_2) &= \frac{N^2}{Z}\iiiint d\rr_3\cdots d\rr_N \prod_{i<j} \Theta(r_{ij}-\sigma)
\end{align}
\begin{align}
  g^{(2)}(\rr_1,\rr_2) &=
  \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
  \\
  &= \frac{V^2}{Z}\iiiint d\rr_3\cdots d\rr_N \prod_{i<j}
  \Theta(r_{ij}-\sigma)
  \label{eq:g2-integral-defn}
\end{align}

\begin{figure}
  \includegraphics[width=3cm]{figs/dadz-3-2.pdf}
  \includegraphics[width=3cm]{figs/dadz-3-3.pdf}
  \includegraphics[width=3cm]{figs/dadz-2-2.pdf}
  \includegraphics[width=3cm]{figs/dadz-2-3.pdf}
  %%\includegraphics[width=3cm]{figs/dadz-4-3}
  \caption{Plot of $\frac{da_1}{dz}$ near a hard
    wall. For what $\Phi$?}\label{fig:dadz}
\end{figure}

At this point we consider the effect of increasing the diameter
$\sigma$ of all our spheres.  This will reduce the number of available
states, thus reducing the partition function $Z$, which we will
ultimately relate to a change in free energy.
\begin{align}
  Z(\sigma+\delta) &= \iiiint d\rr_1\cdots d\rr_N \prod_{i<j} \Theta(r_{ij}-\sigma-\delta)
  \\
  &= \iiiint d\rr_1\cdots d\rr_N \prod_{i<j} \Theta(r_{ij}-\sigma)(1-\Theta(\sigma+\delta-r_{ij}))
\end{align}
I have written $Z(\sigma+\delta)$ in such a way that it is clear that
it is equal to $Z(\sigma)$ plus a small quantity, provided $\delta$ is
small.  We will now investigate that small quantity, which we will
write as a power series in $\delta$.  We begin by expanding the
product of $(1-\Theta(\sigma+\delta-r_{ij}))$:
\begin{multline}
    \prod_{i<j}(1-\Theta(\sigma+\delta-r_{ij})) =
    1 - \sum_{i<j} \Theta(\sigma+\delta-r_{ij}) \\
    + \mathop{\mathop{\sum_{i<j}}_{k<}}_{kl \ne ij}
    \Theta(\sigma+\delta-r_{ij}) \Theta(\sigma+\delta-r_{kl})
    \\ + \cdots.
\end{multline}
Given this expansion, we can readily expand the partition function:
\begin{multline}
  Z(\sigma+\delta) = Z(\sigma) \\ -
  \sum_{k<l} \iiiint \Theta(\sigma+\delta-r_{kl}) \prod_{i<j}
  \Theta(r_{ij}-\sigma) \\
  + \mathop{\mathop{\sum_{k<l}}_{m<n}}_{kl\ne mn}
  \iiiint \Theta(\sigma+\delta-r_{kl}) \Theta(\sigma+\delta-r_{mn}) \prod_{i<j}
  \Theta(r_{ij}-\sigma) \\
  + \cdots.\label{eq:z-power-series}
\end{multline}
Our challenge now is to manipulate this into forms involving $g(r)$,
which we can then simplify using a Taylor expansion around
$r=\sigma$.  We will begin by taking advantage of the fact that the
summations in Eq.~\ref{eq:z-power-series} are mostly of identical
terms, since the integrals are over all coordinates.
\begin{multline}
  Z(\sigma)-Z(\sigma+\delta) =
  N^2\iiiint \Theta(\sigma+\delta-r_{12})
  \prod_{i<j} \Theta(r_{ij}-\sigma) \\
  - \frac{N^3}{2}\iiiint
  \Theta(\sigma+\delta-r_{12})
  \Theta(\sigma+\delta-r_{34})
  \prod_{i<j} \Theta(r_{ij}-\sigma) \\
  + \cdots \label{eq:z-expansion}
\end{multline}
At this point we are ready to connect with $g(r)$, which we can do by
considering an integral of the pair distribution function, from
Eq.~\ref{eq:g2-integral-defn}.
\begin{multline}
  \int_{\sigma}^{\sigma+\delta} g(r) d\rr\\
  \shoveleft{\quad= \int g^{(2)}(\rr_1,\rr_2)
  \Theta(\sigma+\delta-r_{12})d\rr_2} \\
  \shoveleft{\quad= \frac{1}{V}\int g^{(2)}(\rr_1,\rr_2)
  \Theta(\sigma+\delta-r_{12})d\rr_1d\rr_2} \\
  \shoveleft{\quad= \frac{V}{Z}\iiiint d\rr_1\cdots d\rr_N
  \Theta(\sigma+\delta-r_{12})}\\
  \prod_{i<j} \Theta(r_{ij}-\sigma)
  \label{eq:g-shell-integral}
\end{multline}
Looking back at Eq.~\ref{eq:z-expansion}, we can see that the first
term is easily related to Eq.~\ref{eq:g-shell-integral}.  The second
term has two shell $\delta$ functions, and is thus a bit trickier.
Fortunately, in a very large system, $\rr_3$ and $\rr_4$ are
vanishingly seldom near $\rr_1$ or $\rr_2$, so we can write the second
term as a product.
\begin{multline}
  \frac{Z(\sigma)-Z(\sigma+\delta)}{Z(\sigma)}\\
  \shoveleft{\quad= nN \int_{\sigma}^{\sigma+\delta} g(r) d\rr }
  - \frac{n^2N}{2} \left(\int_{\sigma}^{\sigma+\delta} g(r)
  d\rr\right)^2
  \\
  + \frac{n^3N}{6} \left(\int_{\sigma}^{\sigma+\delta} g(r)
  d\rr\right)^3 + \cdots
  \label{eq:zdiff-g}
\end{multline}
Now let us look at how this integral relates to $g(r)$ evaluated at
contact.  I'll skip a few steps doing this integral...
\begin{equation}
   \int_{\sigma}^{\sigma+\delta} g(r)
  d\rr = 4\pi\sigma^2\left(g_\sigma\delta + \left(\tfrac12g_\sigma' +
  \tfrac1{\sigma}\right) + \cdots \right)
\end{equation}
Putting this together with the above we find
\begin{multline}
  \frac{Z(\sigma)-Z(\sigma+\delta)}{Z(\sigma)} =
  4\pi\sigma^2 n g_\sigma \delta +
  4\pi\sigma^2 n \left(\tfrac{g_\sigma'}{2} + \tfrac{2}{\sigma}\right)
  \delta^2 \\
  - \tfrac{1}{2}\left(4\pi\sigma^2ng_\sigma\right)^2\delta^2 + \mathcal{O}(\delta^3)
\end{multline}
But by considering a Taylor expansion, we can also see that
\begin{multline}
  \frac{Z(\sigma)-Z(\sigma+\delta)}{Z(\sigma)} =
  -\frac{\partial Z}{\partial \sigma} \delta
  -\frac12 \frac{\partial^2 Z}{\partial \sigma^2} \delta^2  + \mathcal{O}(\delta^3)
\end{multline}
from which we can extract relationships for the derivatives of $Z$
with respect to $\sigma$.  Which if we knew what those derivatives
were, would tell us about $g_\sigma'$ and $g_\sigma$, which is what we
are after here.  We find this out by considering the Helmholtz free
energy.
\begin{equation}
  F = -kT \ln Z
\end{equation}
The derivatives are then given by
\begin{equation}
  \frac{\partial F}{\partial \sigma} = -\frac{kT}{Z} \frac{\partial
    Z}{\partial \sigma}
\end{equation}
and
\begin{equation}
  \frac{\partial^2 F}{\partial \sigma^2} =
  \frac{1}{kT}\left(\frac{\partial F}{\partial \sigma}\right)^2
  -\frac{kT}{Z} \frac{\partial^2
    Z}{\partial \sigma^2}.
\end{equation}
With a bunch more work, we find out that
\begin{equation}
  g_\sigma' = \frac{\frac{\partial^2 F}{\partial
      \sigma^2}}{4\pi\sigma^2nkT}
  - \frac{2g_\sigma}{\sigma}
\end{equation}
except that there is a mistake in here.  I can tell because $F$ is
extensive and $g_\sigma'$ is not.  This tracks back to
Eq.~\ref{eq:zdiff-g}, which is correct in this \LaTeX, but is wrong on
my scratch paper.
